---
title: "EPD Prescribing Dashboard"
author: "Marco Sorbona"
date: "April 2026"
output: 
  flexdashboard::flex_dashboard:
    orientation: rows
    vertical_layout: fill
    theme: flatly
    self_contained: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(flexdashboard)
library(tidyverse)
library(plotly)
library(DT)
library(arrow)

# Load the curated data
curated <- read_parquet("data/clean/epd_curated.parquet")

# Define ALL variables here (before sidebar tries to use them)
total_items <- sum(curated$items, na.rm = TRUE)
total_cost <- sum(curated$actual_cost, na.rm = TRUE)
unique_practices <- n_distinct(curated$practice_code)
unique_drugs <- n_distinct(curated$bnf_presentation_code)
date_min <- min(curated$year_month, na.rm = TRUE)
date_max <- max(curated$year_month, na.rm = TRUE)
corruption_count <- sum(!is.na(curated$corruption_flag))
```

# Overview {.sidebar}

## Data Summary

------------------------------------------------------------------------

### Total Items

`r format(total_items, big.mark = ",")`

------------------------------------------------------------------------

### Total Cost

`r paste0("£", format(round(total_cost, 0), big.mark = ","))`

------------------------------------------------------------------------

### Unique Practices

`r format(unique_practices, big.mark = ",")`

------------------------------------------------------------------------

### Unique Drugs

`r format(unique_drugs, big.mark = ",")`

### Date Range

`r date_min` to `r date_max`

**Months:** `r n_distinct(curated$year_month)`

### Data Quality Note

`r if(corruption_count > 0) paste0("⚠️ **Corruption detected:** ", corruption_count, " rows flagged") else paste0("✅ **No corruption flags** in this sample")`

*Note: API limited to 10,000 rows per month*

⚠️ **Disclaimer:** Dashboard based on 40,000 row sample only

### Data Quality Discovery

**2023 API Limitation:** November and December 2023 records have missing drug names (20,000 rows, £913k). Later months (2025-2026) contain complete drug names.

*This finding is documented in the quality report.*

# Top Drugs {.column}

### Top 10 Drugs by Total Cost

```{r}
top_drugs <- curated |> 
  group_by(bnf_presentation_name) |> 
  summarise(total_cost = sum(actual_cost, na.rm = TRUE)) |> 
  arrange(desc(total_cost)) |> 
  head(10)

ggplot(top_drugs, aes(x = reorder(bnf_presentation_name, total_cost), 
                       y = total_cost, 
                       fill = total_cost)) +
  geom_col() +
  coord_flip() +
  scale_fill_gradient(low = "lightblue", high = "darkblue") +
  labs(x = "", y = "Total Cost (£)", 
       title = "Top 10 Drugs by Cost") +
  theme_minimal() +
  theme(legend.position = "none")

```

### Data Quality: Missing Drug Names by Month

```{r}
missing_by_month <- curated|>
  group_by(year_month)|>
  summarise(
    missing_drug_names = sum(is.na(bnf_presentation_name)),
    total_rows = n(),
    pct_missing = 100 * missing_drug_names / total_rows
  )

# Create dynamic colour palette
colour_palette <- missing_by_month |>
  mutate(colour = ifelse(pct_missing == 100, "#D62728", "#2CA02C")) |>
  select(year_month, colour) |>
  deframe()

ggplot(missing_by_month, aes(x = year_month, y = pct_missing, fill = year_month)) +
  geom_col() +
  geom_text(aes(label = paste0(round(pct_missing, 0), "%")), vjust = -0.5, size = 3) +
  scale_fill_manual(values = colour_palette) +
  labs(x = "Month", y = "Missing Drug Names (%)", 
       title = "Data Quality: Missing Drug Names by Month",
       subtitle = paste0("Months with 100% missing: ", 
                         paste(missing_by_month |> filter(pct_missing == 100) |> pull(year_month), collapse = ", "),
                         " | 0% missing: ",
                         paste(missing_by_month |> filter(pct_missing == 0) |> pull(year_month), collapse = ", "))) +
  theme_minimal() +
  theme(legend.position = "none")
```

**Insight:** The API endpoint for 2023 data (November and December) returned zero drug names - affecting 20,000 rows and £913k in prescribing costs. Later months (2025-2026) contain complete drug names, suggesting a data quality improvement over time or a limitation in the historical API endpoint.

# Regional Analysis {.column}

### Prescribing by Region (Time Series)

```{r}
region_summary <- curated |> 
  group_by(regional_office_name, year_month) |> 
  summarise(
    total_cost = sum(actual_cost, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(region_summary, aes(x = year_month, 
                            y = total_cost, 
                            color = regional_office_name,
                            group = regional_office_name)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.3) +
  labs(x = "Month", y = "Total Cost (£)", 
       title = "Prescribing Cost by Region Over Time",
       color = "Region") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Regional Cost Share

```{r}
region_share <- curated |> 
  group_by(regional_office_name) |> 
  summarise(total_cost = sum(actual_cost, na.rm = TRUE)) |> 
  arrange(desc(total_cost))

ggplot(region_share, aes(x = reorder(regional_office_name, total_cost), 
                          y = total_cost, 
                          fill = regional_office_name)) +
  geom_col() +
  coord_flip() +
  labs(x = "", y = "Total Cost (£)", 
       title = "Total Cost by Region") +
  theme_minimal() +
  theme(legend.position = "none")
```

# Latest Month Analysis {.column}

### Latest Month Snapshot

```{r}
# Find the latest month in the data
latest_month <- curated |>
  filter(!is.na(bnf_presentation_name)) |>
  arrange(desc(year_month)) |>
  slice(1) |>
  pull(year_month)

latest_month_data <- curated |> 
  filter(year_month == latest_month)

if(nrow(latest_month_data) > 0) {
  latest_items <- sum(latest_month_data$items, na.rm = TRUE)
  latest_cost <- sum(latest_month_data$actual_cost, na.rm = TRUE)
  latest_drugs <- n_distinct(latest_month_data$bnf_presentation_name, na.rm = TRUE)
  
  cat("**Month:** ", latest_month, "\n\n")
  cat("**Total Items:**", format(latest_items, big.mark = ","), "\n\n")
  cat("**Total Cost:** £", format(round(latest_cost, 0), big.mark = ","), "\n\n")
  cat("**Unique Drugs:**", format(latest_drugs, big.mark = ","), "\n\n")
  cat("**Rows:**", nrow(latest_month_data))
} else {
  cat("No complete month data available")
}
```

### Month-by-Month Comparison

```{r}
monthly_summary <- curated |> 
  group_by(year_month) |> 
  summarise(
    total_items = sum(items, na.rm = TRUE),
    total_cost = sum(actual_cost, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(monthly_summary, aes(x = year_month)) +
  geom_col(aes(y = total_items), fill = "steelblue", alpha = 0.7) +
  geom_line(aes(y = total_cost / 10), group = 1, color = "red", linewidth = 0.8) +
  scale_y_continuous(
    name = "Total Items",
    sec.axis = sec_axis(~ . * 10, name = "Total Cost (£)")
  ) +
  labs(x = "Month", title = "Monthly Prescribing Volume and Cost") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# Key Insights {.data}

### Summary of Findings

```{r}
# Create an insights table
insights <- data.frame(
  Finding = c(
    "Missing drug names (2023 API data)",
    "Inconsistent date formats",
    "Later months have complete data",
    "Regional variation observed (in sample)",
    "Sample limitations"
  ),
  Insight = c(
    "API endpoint for 2023 data lacks drug names – data quality issue",
    "API returns different date formats for different years",
    "Data quality improved by 2025 – 0% missing drug names in sample",
    "London and South East had highest costs in this sample",
    "Findings based on 40k random sample (10k/month)"
  ),
  Qualification = c(
    "✅ Affects ALL users of 2023 API data",
    "✅ Affects ALL users of the API",
    "✅ Affects ALL users of 2025+ API data",
    "⚠️ Sample only – not generalisable",
    "⚠️ Do not draw national conclusions"
  )
)

datatable(insights, 
  options = list(pageLength = 10, scrollX = TRUE),
  rownames = FALSE,
  caption = "Key Findings from EPD Curation (40,000 row sample)"
)
```

# Raw Data  {.data}

### Curated Dataset Preview

```{r}
datatable(curated |> 
  select(year_month, regional_office_name, practice_name_raw, 
         bnf_presentation_name, items, actual_cost) |> 
  head(100),
  options = list(scrollX = TRUE, pageLength = 10),
  rownames = FALSE)
```

