---
title: "English Prescribing Dataset (EPD) Curation Report"
author: "Marco Sorbona"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(arrow)
library(dplyr)
library(ggplot2)
library(DT)
library(tidyverse)
```

## Executive Summary

This report documents the curation of the **English Prescribing Dataset (EPD) with SNOMED Code**, a monthly NHS dataset capturing all primary care prescriptions dispensed in England.

**Data scale**: 7 GB per month, 17M+ rows monthly

**Curation approach**: CKAN API with R data frame processing (sample: 10,000 rows per month)

**Memory usage**: \< 2 GB

```{r summary-stats, echo=FALSE}
# Load curated data (use relative path)
curated <- read_parquet("C:/Users/Marco/Documents/Repos_Data_Science_with_R/nhs-prescribing-data-curation/data/clean/epd_curated.parquet")

# Generate summary stats dynamically
total_rows <- nrow(curated)
unique_months <- n_distinct(curated$year_month)
min_month <- min(curated$year_month, na.rm = TRUE)
max_month <- max(curated$year_month, na.rm = TRUE)
total_items <- sum(curated$items, na.rm = TRUE)
total_cost <- sum(curated$actual_cost, na.rm = TRUE)

cat("**Sample processed**:", total_rows, "rows across", unique_months, "months (", min_month, "to", max_month, ")\n")
```

## Data Source & Attribution

**Source**: NHS Business Services Authority. "English Prescribing Dataset (EPD) with SNOMED Code" NHSBSA Open Data. Accessed `r Sys.Date()`. URL: <https://opendata.nhsbsa.net/dataset/english-prescribing-dataset-epd-with-snomed-code>

**Access Method**: CKAN API (`https://opendata.nhsbsa.net/api/3/action/datastore_search`)

**Licence**: Open Government Licence v3.0

**Copyright**: NHSBSA Copyright 2025

## Key Data Quality Discoveries

### 1. Missing Drug Names in API Data

**Challenge**: The API endpoint for certain years returned zero `bnf_presentation_name` values.

```{r missing-impact, echo=FALSE}
# Dynamically detect which months/years have missing drug names
months_with_missing <- curated |>
  filter(is.na(bnf_presentation_name)) |>
  distinct(year_month) |>
  pull(year_month)

# Extract unique years from affected months
years_affected <- unique(substr(months_with_missing, 1, 4))

if(length(months_with_missing) > 0) {
  missing_data <- curated |>
    filter(year_month %in% months_with_missing) |>
    summarise(
      rows = n(),
      items = sum(items, na.rm = TRUE),
      cost = sum(actual_cost, na.rm = TRUE)
    )
  
  # Dynamic title based on affected years
  if(length(years_affected) == 1) {
    cat("**Affected year(s):** ", years_affected, "\n\n", sep = "")
  } else {
    cat("**Affected years:** ", paste(years_affected, collapse = ", "), "\n\n", sep = "")
  }
  
  cat("- **", missing_data$rows, " rows** affected\n", sep = "")
  cat("- **", format(missing_data$items, big.mark = ","), " items**\n", sep = "")
  cat("- **£", format(round(missing_data$cost, 0), big.mark = ","), " in prescribing costs** cannot be attributed to specific drugs\n", sep = "")
} else {
  cat("**No missing drug names detected** in this sample.\n")
}
```

**Solution**: Documented limitation; flagged affected rows with `corruption_flag`; recommend excluding affected months for drug-level analysis.

```{r missing-table, echo=FALSE}
missing_by_month <- curated |>
  group_by(year_month) |>
  summarise(
    rows = n(),
    items = sum(items, na.rm = TRUE),
    cost = sum(actual_cost, na.rm = TRUE),
    missing_drug_names = sum(is.na(bnf_presentation_name)),
    pct_missing = 100 * missing_drug_names / rows
  ) |>
  mutate(
    drug_names_present = case_when(
      pct_missing == 100 ~ "❌ ALL missing",
      pct_missing == 0 ~ "✅ ALL present",
      TRUE ~ paste0(100 - pct_missing, "% present")
    )
  )

knitr::kable(missing_by_month |> select(year_month, rows, items, cost, drug_names_present),
  caption = "Missing Drug Names by Month")
```

### 2. Inconsistent Date Formats Across Years

**Challenge**: API returned different date formats for different years (observed in the data).

**Solution**: Standardised using `case_when()` before time series analysis.

### 3. Serious Shortage Protocols (SSPs)

**Challenge**: No flag for SSP claims; dispensed drug ≠ prescribed drug.

**Limitation**: Cannot identify SSP-impacted prescriptions from EPD alone.

**Mitigation**: Documented limitation in quality report and dashboard disclaimer.

### 4. Negative Quantities (Returns)

**Challenge**: Returns to pharmacy appear as negative quantities.

**Decision**: Retain in dataset; documented.

### 5. Missing Lookup Matches

**Challenge**: Practice and BNF codes may not match lookup tables.

**Action**: Left joins with reporting of match rates.

## Quality Metrics

```{r quality-metrics, echo=FALSE}
# Load quality report if it exists
if(file.exists("data/clean/quality_report.rds")) {
  quality <- readRDS("data/clean/quality_report.rds")
  
  cat("### Summary Metrics\n")
  cat("- **Total rows processed:** ", quality$total_rows_after_cleaning, "\n")
  cat("- **Months processed:** ", paste(quality$months_processed, collapse = ", "), "\n")
  cat("- **Date range:** ", quality$date_range$earliest, " to ", quality$date_range$latest, "\n")
  cat("- **Total items:** ", format(quality$final_summary$total_items, big.mark = ","), "\n")
  cat("- **Total cost:** £", format(round(quality$final_summary$total_cost, 0), big.mark = ","), "\n")
  cat("- **Unique practices:** ", format(quality$final_summary$unique_practices, big.mark = ","), "\n")
  cat("- **Unique drugs:** ", format(quality$final_summary$unique_drugs, big.mark = ","), "\n")
  
  if(!is.null(quality$data_quality_issues)) {
    cat("\n### Data Quality Issues Identified\n")
    cat("- **Missing drug names (2023):** ", quality$data_quality_issues$missing_drug_names$affected_rows, " rows affected\n")
    cat("- **Inconsistent date formats:** 2023 vs 2025/2026\n")
  }
} else {
  cat("Run `01_curate_epd_api.R` first to generate quality report.")
}
```

## Key Variables

| Variable | Type | Description |
|------------------------|------------------|------------------------------|
| `year_month` | character | Year and month (format varies by year) |
| `practice_code` | character | NHS GP practice code (ODS code) |
| `bnf_presentation_code` | character | 15-digit BNF drug code |
| `bnf_presentation_name` | character | Drug name (⚠️ missing for years without data) |
| `items` | numeric | Number of prescription items |
| `actual_cost` | numeric | Reimbursement cost in GBP |
| `regional_office_name` | character | NHS England region |
| `corruption_flag` | character | Flags rows with known data issues |
| `date` | Date | Standardised date for time series |

## Dynamic Sample Observations

```{r dynamic-insights, echo=FALSE}
# Generate top drugs dynamically
top_drugs <- curated |>
  filter(!is.na(bnf_presentation_name)) |>
  group_by(bnf_presentation_name) |>
  summarise(total_cost = sum(actual_cost, na.rm = TRUE)) |>
  arrange(desc(total_cost)) |>
  head(5)

cat("### Top 5 Drugs in This Sample\n")
print(top_drugs)

# Regional summary
region_summary <- curated |>
  group_by(regional_office_name) |>
  summarise(total_cost = sum(actual_cost, na.rm = TRUE)) |>
  arrange(desc(total_cost))

cat("\n### Regional Cost Distribution\n")
print(region_summary)
```

## Limitations (Official NHSBSA Caveats)

1.  **PGD exclusions**: Pharmacy First, contraception, smoking cessation not captured
2.  **Dental prescribing**: Excluded entirely
3.  **Hospital site codes**: Excluded (only cost centre codes included)
4.  **SSPs**: Cannot be identified from EPD alone
5.  **Code versioning**: BNF/SNOMED codes may change over time

## Project Limitations

| Limitation | Impact | Mitigation |
|---------------------------|------------------|---------------------------|
| API limit (10k rows/month) | Findings based on sample | Clear disclaimers |
| Missing drug names (some years) | Cannot analyse those months | Flagged in quality report |
| Limited months | Restricted time series | Documented; pipeline scales |

## Data Quality Issue: Missing Drug Names

### Discovery

The API endpoint for certain years returned zero drug names.

### Impact Assessment

-   Drug-level analysis not possible for affected months
-   Regional and monthly totals remain valid

### Proposed Resolution (Production Context)

| Step | Action                                     |
|------|--------------------------------------------|
| 1    | Download full CSV files from NHSBSA portal |
| 2    | Use CSV pipeline instead of API            |
| 3    | Join with appropriate BNF lookup           |

### Decision for This Portfolio

Documented as a data quality discovery rather than resolved, to demonstrate issue identification skills.

## Usage Notes

To reproduce this curation:

1.  **Clone** this repository
2.  **Ensure lookup tables** are in `data/lookup/`
3.  **Run** the curation pipeline: `Rscript R/01_curate_epd_api.R`
4.  **Render** this report: `quarto render curation_report.qmd`
5.  **Render** the dashboard: `rmarkdown::render("epd_dashboard.Rmd")`

## References

-   NHSBSA. English Prescribing Dataset Release Guidance.
-   Open Government Licence v3.0.
-   BNF Code Information - Current Year. NHSBSA Open Data Portal.

## Data Source

NHS Business Services Authority. "English Prescribing Dataset (EPD) with SNOMED Code" OGL v3.0. NHSBSA Copyright 2025.

## Accessible Dashboard

An interactive version of this analysis is available at:

[Live Dashboad](https://marcosorbona.github.io/nhs-prescribing-data-curation/epd_dashboard.html)
